TSELM: Target Speaker Extraction using Discrete Tokens and Language Models

### 2024-09-12 src/paper.tex
update

\begin{IEEEkeywords}
target speaker extraction, speech separation, language models, audio discretization, Self-Supervised Models
\end{IEEEkeywords}
\begin{figure*}[t]
    \centering
    \includegraphics[width=0.77\textwidth]{assets/model.pdf}
    \caption{Overview of TSELM.}
    \label{model}
    \vspace{-12pt}
    \end{figure*}

@@

The modeling of discretization of audio has gained significant attention with the advancement of language models (LMs) in modeling texts \cite{l1, l2, l3, l4, l5}. Researchers are trying to convert audio into discrete tokens and leverage LMs to 
model them, thereby simplifying audio generation tasks by transforming complex regression problems into classification tasks \cite{dasb}.
LMs have demonstrated its scalability and unification of different modalities in audio synthesis tasks \cite{audio_gen,deshmukh2023pengi,chen2023lauragpt}. It has also achieved superior performances compared with discriminative models in
tasks such as speech enhancement \cite{selm,mask_sr} and blind speech separation \cite{dasb,tokensplit}.

@@

This approach is inspired by WavLM \cite{wavlm}, which trains the model by overlapping clean speech with an interfering signal covering less than 50\% of the speech, using the first utterance as the primary speaker. This allows WavLM to focus on producing target speaker dependent embeddings. Our experiments demonstrated that this concatenation strategy significantly enhances the model's performance by guiding it to prioritize the target speaker's information.
\subsection{Modeling}

@@

\subsubsection{Attention Embedding}
After obtaining the discrete tensor \(\bm{d}\) with shape \(6 \times T\), we use 6
learnable embedding tables each with \(K\) entires to embed the 6 layers 
respectively, each resulting in a tensor of shape \(T \times E\). 
After embedding, we follow the same recipe as in \cite{dasb} to 
aggregate the tensor by using attention mechanism to sum all the 6 tensors. This summation 
keeps the information of each layer while reducing the 
dimension of layers. After attention embedding, we obtain reference embedding \(E_r\) and 
mixture embedding \(E_m\).

@@

It is shown that metrics like PESQ, SI-SNR, STOI do not accurately reflect the speech quality of 
vocoder outputs due to the fact that the vocoder output does not focus strictly on frame alignment
\cite{tokensplit,selm}. We use DNSMOS \cite{dnsmos} to measure the speech quality, and the differential 
Word Error 
Rate (dWER) \cite{dwer} using the base model of Whisper \cite{whisper} to measure the speech intelligibility. For speaker similarity, we use the \texttt{ResNet221\_LM}
from the public WeSpeaker \cite{wespeaker} to calculate the embedding cosine speaker similarity.

@@

We observe that TSELM-l outperforms Spex+ in terms of DNSMOS scores, indicating better speech quality, but performs slightly worse in dWER, suggesting lower speech intelligibility. One potential reason for this could be the discretization process applied to the mixed speech. Our Kmeans algorithm is trained on clean speech rather than mixed speech, which is advantageous for speech enhancement as it likely aids in denoising. However, when applied to mixed speech, this discretization might lead to the model focusing on the wrong speaker, as it may only retain dominant speaker information, causing a reduction in intelligibility.
This hypothesis is supported by the results of TSELM-L-Hybrid, where continuous embeddings from the mixture speech were used without discretization, achieving dWER scores better than Spex+ on WSJ0-2mix. 
% Another contributing factor could be the limitations of our current language model (LM). The encoder-only LM achieves around 55\% accuracy for token prediction on Libri2Mix testset, and we believe using more advanced models, such as auto-regressive or masking-based LMs, could lead to improved performance in future iterations.

@@

In Table \ref{hubert_wavlm_6}, we compare the performance of HuBERT and WavLM as SSL models and examine the effects of using either one or multiple layers for discretization. Our findings indicate that using HuBERT as the SSL model results in better DNSMOS scores but much worse dWER compared to our WavLM baseline. The improved DNSMOS scores likely stem from the vocoder performance, yet they do not adequately reflect speech intelligibility, which is crucial for speech separation tasks.
The poorer dWER scores observed with HuBERT may be attributed to its training on clean speech, which might not equip it to capture the complexity and richness of mixed speech. Moreover, our results from WavLM-L6 suggest that when performing speech separation, discretizing across multiple layers provides better results than relying on a single layer.  This might be because that using multiple layer outputs can better tolerate errors compared to using just one layer output.

>>>>

\begin{IEEEkeywords}
target speaker extraction, speech separation, language models, audio discretization, self-supervised models
\end{IEEEkeywords}
\begin{figure*}[t]
    \centering
    \includegraphics[width=0.77\textwidth]{assets/model.pdf}
    \caption{Overview of TSELM.}
    \label{model}
    \vspace{-12pt}
    \end{figure*}

@@

The modeling of discretization of audio has gained significant attention with the advancement of language models (LMs) in modeling texts \cite{l1, l2, l3, l4, l5}. Researchers are trying to convert audio into discrete tokens and leverage LMs to 
model them, thereby simplifying audio generation tasks by transforming complex regression problems into classification tasks \cite{dasb}.
LMs have demonstrated scalability and unification of different modalities in audio synthesis tasks \cite{audio_gen,deshmukh2023pengi,chen2023lauragpt}. It has also achieved superior performances compared with discriminative models in
tasks such as speech enhancement \cite{selm,mask_sr} and blind speech separation \cite{tokensplit}.

@@

This approach is inspired by WavLM \cite{wavlm}, which trains the model by overlapping clean speech with an interfering signal covering less than 50\% of the speech, using the first utterance as the primary speaker. This allows WavLM to focus on producing target speaker dependent embeddings. Our experiments demonstrate that this concatenation strategy significantly enhances the model's performance by guiding it to prioritize the target speaker's information.
\subsection{Modeling}

@@

\subsubsection{Attention Embedding}
After obtaining the discrete tensor \(\bm{d}\) with shape \(6 \times T\), we use 6
learnable embedding tables each with \(K\) entires to embed the 6 layers 
respectively, each resulting in a tensor of shape \(T \times E\). 
After embedding, we follow the same recipe as in \cite{dasb} to 
aggregate the tensor by using attention mechanism to sum all the 6 tensors. This summation 
keeps the information of each layer while reducing the 
dimension of layers and system complexity. After attention embedding, we obtain reference embedding \(E_r\) and 
mixture embedding \(E_m\).

@@

It is shown that metrics like PESQ, SI-SNR, STOI do not accurately reflect the speech quality of 
vocoder outputs due to the fact that the vocoder output does not focus strictly on frame alignment
\cite{tokensplit,selm}. Instead, we use DNSMOS \cite{dnsmos} to measure the speech quality, and the differential 
Word Error 
Rate (dWER) \cite{dwer} using the base model of Whisper \cite{whisper} to measure the speech intelligibility. For speaker similarity, we use the \texttt{ResNet221\_LM}
from the public WeSpeaker \cite{wespeaker} to calculate the embedding cosine speaker similarity.

@@

We observe that TSELM-L outperforms Spex+ in terms of DNSMOS scores, indicating better speech quality, but performs slightly worse in dWER, suggesting lower speech intelligibility. One potential reason for this could be the discretization process applied to the mixed speech. Our Kmeans algorithm is trained on clean speech rather than mixed speech, which is advantageous for speech enhancement as it likely aids in denoising. However, when applied to mixed speech, this discretization might lead to the model focusing on the wrong speaker, as it may only retain dominant speaker information, causing a reduction in intelligibility.
This hypothesis is supported by the results of TSELM-L-Hybrid, where continuous embeddings from the mixture speech were used without discretization, achieving dWER scores better than Spex+ on WSJ0-2mix. 
% Another contributing factor could be the limitations of our current language model (LM). The encoder-only LM achieves around 55\% accuracy for token prediction on Libri2Mix testset, and we believe using more advanced models, such as auto-regressive or masking-based LMs, could lead to improved performance in future iterations.

@@

In Table \ref{hubert_wavlm_6}, we compare the performance of HuBERT and WavLM as SSL models and examine the effects of using either one or multiple layers for discretization. Our findings indicate that using HuBERT as the SSL model results in better DNSMOS scores but much worse dWER compared to our WavLM baseline. The improved DNSMOS scores likely stem from the vocoder performance, yet they do not adequately reflect speech intelligibility, which is crucial for speech separation tasks.
The poorer dWER scores observed with this HuBERT model may be attributed to its training on clean speech, which might not equip it to capture the complexity and richness of mixed speech. Moreover, our results from WavLM-L6 suggest that when performing speech separation, discretizing across multiple layers provides better results than relying on a single layer.  This might be because that using multiple layer outputs can better tolerate errors compared to using just one layer output.
